---
title: "MICOM time series comparison"
author: "Laurie and Casey"
format: html
editor: visual
---

# Comparison of MICOM-predicted growth rates and measured changes in abundance from time series data

Load in libraries

```{r}
library(cowplot) # to plot grid
library(glue) # to format and interpolate a string
library(lme4) # for linear modeling 
library(tidyverse) 
library(ggh4x) # ggplot2 extension, includes nested facets
library(broom)
```

## Load in actual clr-transformed relative abundance changes (output of calculate_actual_growth_rates_genus_clr.py)

Convert df from wide to long and make sure columns match for joining real and sim data later

```{r}
#create long df with one subject
# actual_clr_wide_F01 <- read_csv("../data/actual_growth_rates_genus_clr/F01_clr_actual_growth_rates_by_genus.csv", show_col_types = FALSE)
# 
# actual_clr_long_F01 <- actual_clr_wide_F01 %>% 
#   pivot_longer(!Genus, names_to = "sample_id", values_to = "clr_change_abund") %>% 
#   rename(taxon = Genus) %>% 
#   mutate(taxon = str_sub(taxon, 4, -1)) %>% 
#   mutate(sample_id = as.numeric(sample_id)) %>% 
#   # add a column for subject_id
#   cbind(subject_id='F01')

```

```{r}
# create long df with all subjects included, a column added for subject_id, and all 3 subjects combined
#list of subject ids
subject_ids <- c("F01", "M01", "M02")

# Use map_dfr (from purrr package) to read, transform, and combine dataframes
actual_clr_long_combined <- map_dfr(subject_ids, function(subject) {
  
  # Read each subject's file
  df_wide <- read_csv(paste0("../data/actual_growth_rates_genus_clr/", subject, "_clr_actual_growth_rates_by_genus.csv"), 
                      show_col_types = FALSE)
  
  # Transform to long format and clean columns
  df_long <- df_wide %>% 
    pivot_longer(cols = !Genus, names_to = "sample_id", values_to = "clr_change_abund") %>% 
    rename(taxon = Genus) %>% 
    mutate(taxon = str_sub(taxon, 4, -1),
           sample_id = as.numeric(sample_id),
           subject_id = subject)  # add subject_id column directly
  
  return(df_long)
})

# Verify combined dataframe
head(actual_clr_long_combined)
```

## Load in the MICOM-predicted growth data (output of simulate_growth_rates.py script)

```{r echo=FALSE, include=FALSE, message=FALSE}

#grab list of all .zip files with the pattern we want then remove .zip from the end
zip_list <- Sys.glob("../data/growth_rates/growth*.zip")
folder_list <- gsub(basename(zip_list), pattern = ".zip", replacement = "")

#establish pattern for file recognition in for loop
growth_file_pattern <- "../data/growth_rates/{folder_name}/growth_rates.csv"

# populate an empty list with dataframes 
growth_df_list <- list()

for (folder_name in folder_list) {
    tmp_fp <- glue(growth_file_pattern)
    tmp_df <- read_csv(tmp_fp)
    tmp_df$folder_name <- folder_name
    #separate folder_name by "_" into new columns
    tmp_df <- separate(tmp_df, 
                       folder_name, 
                       into = c("micom_step", 
                                "subject_id", 
                                "model_db",
                                "solver", 
                                "diet", 
                                "tradeoff"), 
                       sep = "_", 
                       remove = FALSE #if TRUE, removes original col
                       )
    #create named list (each element named for its corresponding folder_name)
    #naming an element in a list uses [[double brackets]]
    growth_df_list[[folder_name]] <- tmp_df
    #print(head(tmp_df))
}

#create a df that includes all data from all folders in folder_list
growth_df <- bind_rows(growth_df_list)
```

delete the g\_\_ in front of each genus name to match the actual growth rate taxon names

```{r}
micom_rates <- growth_df %>% 
  mutate(taxon = str_sub(taxon, 4, -1)) %>% 
  group_by(folder_name, sample_id)
  #originally had CLR transform, but these growth rates are not compositional
  # mutate(sample_growth_rate_geom_mean = exp(mean(log(growth_rate))),
  #        sample_growth_rate_clr = log(growth_rate/sample_growth_rate_geom_mean))
```

Later, I want to be able to sort by only the most prevalent taxa. Here, we set a cutoff for total fraction of days that a genus must be present to be "prevalent" (e.g. 0.90 is present in at least 90% of time points)

```{r}
PREVALENCE <- 0.90
```

For graphing, filter out prevalent taxa (present in at least 90% of samples).

```{r}
PREVALENCE <- 0.90

prevalent_micom_rates <- micom_rates %>%
  group_by(folder_name) %>% 
  mutate(total_samples = n_distinct(sample_id)) %>% 
  ungroup() %>% 
  group_by(folder_name, taxon) %>%
  mutate(sample_counts = n_distinct(sample_id[!is.na(growth_rate)]),
         prevalence = sample_counts / total_samples) %>%
  ungroup() %>%
  filter(prevalence >= PREVALENCE) %>%
  mutate(normed_epoch_time = (sample_id - min(sample_id)) /
                             (max(sample_id) - min(sample_id))) %>%
  mutate(sample_id = as.numeric(sample_id))

```

## Join Data

Joining actual with only the prevalent micom rates (\>90% of samples) to cut down on number of comparisons

```{r}
sim_real_data <- left_join(prevalent_micom_rates, 
                           actual_clr_long_combined, 
                           by=c("taxon","sample_id", "subject_id")) %>% 
  drop_na(growth_rate, clr_change_abund) %>% 
  mutate(sampling_date = as.POSIXct(sample_id, origin = "1970-01-01", tz = "UTC"))

```

For graphing, want to add two more columns for the onset of illness

-   column called "sick_day" that adds the correct date for each subject's onset of illness (in metadata)

-   column called timing_vs_sickness_onset that is "before_onset" if before "sick_day" and "after_onset" if on or after sick_day

```{r}

# #F01 sick day
# F01_sick_day = as.POSIXct(1202688000, origin = "1970-01-01", tz = "UTC")
# #M01 sick day 
# M01_sick_day = as.POSIXct(1203465600, origin = "1970-01-01", tz = "UTC")
# #M02 sick day
# M02_sick_day = as.POSIXct(1203120000, origin = "1970-01-01", tz = "UTC")
```

```{r}
sim_real_data <- sim_real_data %>% 
  #mutate adds or modifies columns
  #case_when allows you to vectorize multiple "if else" statements 
  mutate(sick_day = case_when(
    subject_id == "F01" ~ 1202688000, 
    subject_id == "M01" ~ 1203465600,
    subject_id == "M02" ~ 1203120000)) %>% 
  mutate(date_vs_onset_illness = case_when(
    sample_id < sick_day ~ "before", 
    sample_id >= sick_day ~ "after"
  ))
```

## 

## Correlation

I want to see the correlation between

-   can I write a for loop to cycle though all of these combinations - top level: folder_name/parameters, further nested by taxon, possibly further nested by before vs. after onset of illness?

-   how can I extract the values that I want from the Spearman test. Do I filter first by significance?

-   Can I average the rho values across all the taxa in a single folder?

-   How can I quantify what parameters are performing better than others

    -   rank by p-value and rho, count instances of each parameter and where they fall in the ranking?

```{r}
growth_F01_agora201_gurobi_wd_08_Akkermansia <- sim_real_data %>% 
  filter(folder_name == "growth_F01_agora201_gurobi_wd_08", taxon == "Akkermansia")

spearman_cor <- cor.test(growth_F01_agora201_gurobi_wd_08_Akkermansia$growth_rate, growth_F01_agora201_gurobi_wd_08_Akkermansia$clr_change_abund, method = c("spearman"))

spearman_cor
```

Using group_by and broom library to cycle through parameter combinations and taxa

```{r}
spearman_results <- sim_real_data %>% 
  #group by folder_name and taxon then perform spearman correlation tests systematically 
  group_by(folder_name, taxon, subject_id, model_db, diet, tradeoff, date_vs_onset_illness) %>% 
  #nest rows into a list-column of dfs, each unique group has its own df with all rows belonging to that group (data column)
  nest() %>% 
  #first filter out groups with non-zero variance (was causing errors in several samples)
  #map() applies a function to each element of a vector (map_dbl return a vector of type dbl)
  mutate(
    var_growth = map_dbl(data, ~ sd(.x$growth_rate, na.rm=TRUE)), 
    var_abund = map_dbl(data, ~sd(.x$clr_change_abund, na.rm=TRUE))
  ) %>% 
  #removes zero variance cases
  filter(var_growth > 0 & var_abund > 0) %>% 
  mutate(
    #map applies a function to each element of a vector
    spearman = map(data, ~ cor.test(.x$growth_rate, 
                                    .x$clr_change_abund, 
                                    method = "spearman", 
                                    #received warning about computing exact p-value with ties
                                    #default to approx. p-values to deal with this
                                    exact = FALSE)),
    #extracts the rho and p-values from spearman with broom::tidy()
    tidy_results = map(spearman, tidy)
  ) %>% 
  #extract tidy results into a new df 
  unnest(tidy_results) %>% 
  select(-data, -spearman, -var_growth, -var_abund)

nrow(spearman_results)
summary(spearman_results)
```

filter results by significance using broom

```{r}
significant_spearman_p <- spearman_results %>% 
  filter(p.value < 0.05)

nrow(significant_spearman_p)
summary(significant_spearman_p)
```

average rho values across all taxa within the same folder

```{r}
avg_rho_by_folder <- spearman_results %>% 
  group_by(folder_name, subject_id, model_db, diet, tradeoff, date_vs_onset_illness) %>% 
  #mean and median rho indicates predictive strength across all taxa
  summarise(mean_rho = mean(estimate, na.rm=TRUE),
            median_rho = median(estimate, na.rm=TRUE), 
            significant_taxa = sum(p.value < 0.05), 
            total_taxa = n()) %>% 
  arrange(desc(mean_rho))

summary(avg_rho_by_folder)
```

quantify which parameters perform better than others

```{r, fig.width=10, fig.height=20}
avg_rho_by_folder %>%
  filter(subject_id == "M02", date_vs_onset_illness == "before") %>% 
  ggplot(aes(x = reorder(folder_name, mean_rho), y = mean_rho, fill = diet)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(x = "Parameter Combination", y = "Mean of Spearman's rho across all genera",
       title = "Ranking MICOM Parameter Sets by Predictive Accuracy (Mean rho)") +
  theme_minimal()

avg_rho_by_folder
```

## Graphing

```{r, fig.width=10, fig.height=20}

clr_actual_vs_micom_rate <- sim_real_data %>% 
  filter(subject_id == "F01", diet == "wd", model_db == "agora201", date_vs_onset_illness ==  "before", tradeoff == "09") %>% 
  ggplot(aes(x = growth_rate, y = clr_change_abund)) + 
  geom_point(alpha=0.3) + 
  geom_smooth(method="lm") + 
  scale_x_continuous(labels=scales::label_number(drop0trailing=TRUE)) + 
  scale_y_continuous(labels=scales::label_number(drop0trailing=TRUE)) + 
  facet_grid2(taxon~tradeoff, scales = "free", independent="x") + 
  theme_bw(base_size=14) + 
  theme(axis.text = element_text(size=10), 
        panel.spacing.x = unit(1, "line"), 
        strip.text.y = element_text(size = 9)) + 
  labs(x = "MICOM-predicted Growth Rate", y = "CLR-transformed Change in Abundance", title = "MICOM Growth Rate Prediction vs. Actual Abundance Changes")

clr_actual_vs_micom_rate
```

# I AM HERE

## Modeling Growth Rate

```{r echo =FALSE}
source("~/Tools/Madi_Code/art_lmer_functions.R")
```

Mixed effect linear modeling (random effect + fixed effect)

```{r}
# micom_growth_rate_model <- lmer(growth_rate ~ abundance + (abundance | taxon),
#                                 data = prevalent_micom_rates)
```

```{r, fig.width=8, fig.height=7.5}
# abundance_vs_growth_rate <- prevalent_micom_rates %>% 
#   ggplot(aes(x = abundance, y = growth_rate)) + 
#   geom_point() +
#   geom_line(aes(y = predict(micom_growth_rate_model)),
#             color="red") +
#   facet_wrap(~taxon,
#              scales="free",
#              ncol=3) +
#   theme_bw(base_size = 15) +
#   labs(x = "Relative Abundance",
#        y = "Growth Rate",
#        title = "Relative Abundance vs. MICOM Growth Rate"
```

## Modeling

\*\*need to come back to mixed linear model for sim_real_growth

```{r}
sim_real_growth_model1 <- lmer(clr_change_abund ~ growth_rate
                               * model_db
                               * diet 
                               * tradeoff 
                               + (1 | taxon)
                               + (1 | subject_id), 
                               data = sim_real_data)

summary(sim_real_growth_model1)
# sim_real_growth_model1 <- lmer(clr_change_abund ~ growth_rate + (1 | taxon),
#                      data = sim_real_data)
# 
# sim_real_growth_model2 <- lmer(log2FC_abund ~ sample_growth_rate_clr +
#                                  (sample_growth_rate_clr | taxon),
#                                data = sim_real_data)
# 
# sim_real_growth_model3 <- lmer(log2FC_abund ~ sample_growth_rate_clr * 
#                                  abundance +
#                                 (abundance | taxon),
#                                data = sim_real_data)

# anova(sim_real_growth_model1, sim_real_growth_model3)
# anova(sim_real_growth_model1, sim_real_growth_model2)
# anova(sim_real_growth_model2, sim_real_growth_model3)
```

Plotting

```{r, fig.width=10, fig.height=20}
growth_rate_vs_fold_change <- sim_real_data %>% 
  ggplot(aes(x = growth_rate, y = log2FC_abund)) +
    geom_point(aes(color= sampling_date)) + 
    geom_smooth(method = "lm") +
    facet_wrap(~taxon, scales = "free", ncol = 3) +
    theme_bw(base_size = 16) +
    labs(x = "Growth Rate",
         y = "Log2 Fold Change",
         title = "MICOM Growth Rate vs. Actual Log2FC Abundance")


growth_rate_vs_fold_change2 <- sim_real_data %>% 
  #filter(diet != "vmhfat") %>% 
  ggplot(aes(x = growth_rate, y = log2FC_abund)) +
  geom_point(alpha=0.3) +
  geom_smooth(method="lm") +
  scale_x_continuous(labels=scales::label_number(drop0trailing=TRUE)) +
  scale_y_continuous(labels=scales::label_number(drop0trailing=TRUE)) +
  facet_grid2(taxon~solver, scales = "free", independent = "x", shrink = TRUE) +
  theme_bw(base_size=14) +
  theme(axis.text = element_text(size=10), 
        panel.spacing.x = unit(1, "line"), 
        strip.text.y = element_text(size = 9))

growth_rate_vs_fold_change2
```
