---
title: "parameter_combo_testing"
author: "Laurie Lyon"
date: "2025-03-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Comparison of MICOM-predicted growth rates and measured changes in relative abundance from time series data

Load in libraries

```{r}
library(cowplot) # to plot grid
library(glue) # to format and interpolate a string
library(lme4) # for linear modeling 
library(tidyverse) 
library(ggh4x) # ggplot2 extension, includes nested facets
library(broom)
```

Load in combined data generated from combine_sim_and_real_data.r (within data folder of micom-time-series project folder)

Choosing file generated from 50% prevalence value cutoff (taxon present in at least half of all sampling timepoints)

```{r}
#sim_real_data <- read_csv("../data/combined_sim_real_data_50.csv")
sim_real_data <- read_csv("../data/combined_sim_real_data.csv")
```

## Measuring Spearman Correlation

Goal: I want to see the correlation between micom predicted growth rates and actual CLR-transformed changes in abundance

Using Spearman Correlation as a nonparametric alternative to measuring monotonic correlation between samples

Plan of attack:

-   can I write a for loop to cycle though all of these combinations - top level: folder_name/parameters, further nested by taxon, possibly further nested by before vs. after onset of illness?

-   how can I extract the values that I want from the Spearman test. Do I filter first by significance?

-   Can I average the rho values across all the taxa in a single folder?

-   How can I quantify what parameters are performing better than others

    -   rank by p-value and rho, count instances of each parameter and where they fall in the ranking?

### Spearman results for all genera, all samples, all parameter combinations

-   further split by date_vs_onset_illness to see if there are differences between the "undisturbed" and "disturbed" gut parameters in this dataset

-   usig group_by and broom to cycle through parameter combinations and taxa

```{r}
spearman_results <- sim_real_data %>% 
  #group by folder_name and taxon then perform spearman correlation tests systematically 
  #group_by(folder_name, taxon, subject_id, model_db, diet, tradeoff, date_vs_onset_illness) %>% 
   group_by(folder_name, taxon, subject_id, model_db, diet, tradeoff) %>% 
  #nest rows into a list-column of dfs, each unique group has its own df with all rows belonging to that group (data column)
  nest() %>% 
  #first filter groups with non-zero variance (was causing errors in several samples)
  #map() applies a function to each element of a vector (map_dbl return a vector of type dbl)
  mutate(
    var_growth = map_dbl(data, ~ sd(.x$growth_rate, na.rm=TRUE)), 
    var_abund = map_dbl(data, ~sd(.x$clr_change_abund, na.rm=TRUE))
  ) %>% 
  #removes zero variance cases
  filter(var_growth > 0 & var_abund > 0) %>% 
  mutate(
    #map applies a function to each element of a vector
    spearman = map(data, ~ cor.test(.x$growth_rate, 
                                    .x$clr_change_abund, 
                                    method = "spearman", 
                                    #received warning about computing exact p-value with ties
                                    #default to approx. p-values to deal with this
                                    exact = FALSE)),
    #extracts the rho and p-values from spearman with broom::tidy()
    tidy_results = map(spearman, tidy)
  ) %>% 
  #extract tidy results into a new df 
  unnest(tidy_results) %>% 
  select(-data, -spearman, -var_growth, -var_abund)

nrow(spearman_results)
# summary(spearman_results)
```

```{r}
# pearson_results <- sim_real_data %>% 
#   #group by folder_name and taxon then perform spearman correlation tests systematically 
#   group_by(folder_name, taxon, subject_id, model_db, diet, tradeoff, date_vs_onset_illness) %>% 
#   #nest rows into a list-column of dfs, each unique group has its own df with all rows belonging to that group (data column)
#   nest() %>% 
#   #first filter groups with non-zero variance (was causing errors in several samples)
#   #map() applies a function to each element of a vector (map_dbl return a vector of type dbl)
#   mutate(
#     var_growth = map_dbl(data, ~ sd(.x$growth_rate, na.rm=TRUE)), 
#     var_abund = map_dbl(data, ~sd(.x$clr_change_abund, na.rm=TRUE))
#   ) %>% 
#   #removes zero variance cases
#   filter(var_growth > 0 & var_abund > 0) %>% 
#   mutate(
#     #map applies a function to each element of a vector
#     pearson = map(data, ~ cor.test(.x$growth_rate, 
#                                     .x$clr_change_abund, 
#                                     method = "pearson", 
#                                     #received warning about computing exact p-value with ties
#                                     #default to approx. p-values to deal with this
#                                     exact = FALSE)),
#     #extracts the rho and p-values from pearson with broom::tidy()
#     tidy_results = map(pearson, tidy)
#   ) %>% 
#   #extract tidy results into a new df 
#   unnest(tidy_results) %>% 
#   select(-data, -pearson, -var_growth, -var_abund)
# 
# nrow(pearson_results)


```

But how many of these are statistically significant?

-   in this example, just over 10% are statistically significant

```{r}
# pearson_significant_p <- pearson_results %>% 
#   filter(p.value < 0.05)
# 
# nrow(pearson_significant_p)

spearman_significant_p <- spearman_results %>% 
  filter(p.value < 0.05)

nrow(spearman_significant_p)

significant_p_percentage <- (nrow(spearman_significant_p)/nrow(spearman_results))*100
```

### Performance Summary - Spearman Correlations

I want a type of summary that tells me how many of these correlations are statistically significant (p \< 0.05), and of these, how many are positive vs. negatively correlated (positive or negative rho value)

```{r}
spearman_significant_p_summary <- spearman_results %>% 
  #add columns for significance and directionality
  mutate(significant = p.value < 0.05, 
         positive = estimate > 0, 
         negative = estimate < 0,
         significant_positive = significant & positive, 
         significant_negative = significant & negative)

```

```{r}
spearman_performance_summary_temporal <- spearman_results %>% 
  #add columns for significance and directionality
  mutate(significant = p.value < 0.05, 
         positive = estimate > 0, 
         negative = estimate < 0,
         significant_positive = significant & positive, 
         significant_negative = significant & negative) %>% 
  group_by(folder_name,
           subject_id,
           model_db,
           diet,
           tradeoff) %>% 
           #date_vs_onset_illness) %>%
  summarise(
    mean_rho = mean(estimate[significant_positive],
                    na.rm = TRUE),
    median_rho = median(estimate[significant_positive],
                        na.rm = TRUE),
    significant_taxa = sum(significant),
    significant_positive_taxa = sum(significant_positive),
    significant_negative_taxa = sum(significant_negative),
    total_taxa = n(),
    percent_significant = (significant_taxa / total_taxa)*100,
    percent_significant_positive = (significant_positive_taxa /
                                    total_taxa)*100,
    percent_significant_negative = (significant_negative_taxa /
                                    total_taxa)*100,
    pos_to_neg_ratio = significant_positive_taxa / (significant_negative_taxa + 1),
    pos_fraction = significant_positive_taxa /
                 (significant_positive_taxa + significant_negative_taxa),
    pos_minus_neg = significant_positive_taxa - significant_negative_taxa
  ) %>%
  #only want positive rho values
  #filter(significant_positive_taxa > 0) %>%
  arrange(desc(pos_minus_neg), desc(percent_significant_positive), desc(median_rho))
```

```{r}
spearman_performance_summary <- spearman_results %>% 
  #add columns for significance and directionality
  mutate(significant = p.value < 0.05, 
         positive = estimate > 0, 
         negative = estimate < 0,
         significant_positive = significant & positive, 
         significant_negative = significant & negative) %>% 
  group_by(folder_name,
           subject_id,
           model_db,
           diet,
           tradeoff) %>%
  summarise(
    mean_rho = mean(estimate[significant_positive],
                    na.rm = TRUE),
    median_rho = median(estimate[significant_positive],
                        na.rm = TRUE),
    significant_taxa = sum(significant),
    significant_positive_taxa = sum(significant_positive),
    significant_negative_taxa = sum(significant_negative),
    total_taxa = n(),
    percent_significant = (significant_taxa / total_taxa)*100,
    percent_significant_positive = (significant_positive_taxa /
                                    total_taxa)*100,
    percent_significant_negative = (significant_negative_taxa /
                                    total_taxa)*100,
    pos_minus_neg = significant_positive_taxa - significant_negative_taxa
  ) %>%
  #only want positive rho values
  #filter(significant_positive_taxa > 0) %>%
  arrange(desc(pos_minus_neg), desc(percent_significant_positive), desc(median_rho))
```

```{r}
# pearson_performance_summary <- pearson_results %>% 
#   #add columns for significance and directionality
#   mutate(significant = p.value < 0.05, 
#          positive = estimate > 0, 
#          negative = estimate < 0,
#          significant_positive = significant & positive, 
#          significant_negative = significant & negative) %>% 
#   group_by(folder_name,
#            subject_id,
#            model_db,
#            diet,
#            tradeoff) %>%
#   summarise(
#     mean_rho = mean(estimate[significant_positive],
#                     na.rm = TRUE),
#     median_rho = median(estimate[significant_positive],
#                         na.rm = TRUE),
#     significant_taxa = sum(significant),
#     significant_positive_taxa = sum(significant_positive),
#     significant_negative_taxa = sum(significant_negative),
#     total_taxa = n(),
#     percent_significant = (significant_taxa / total_taxa)*100,
#     percent_significant_positive = (significant_positive_taxa /
#                                     total_taxa)*100,
#     percent_significant_negative = (significant_negative_taxa /
#                                     total_taxa)*100,
#     pos_minus_neg = significant_positive_taxa - significant_negative_taxa
#   ) %>%
#   #only want positive rho values
#   #filter(significant_positive_taxa > 0) %>%
#   arrange(desc(pos_minus_neg), desc(percent_significant_positive), desc(median_rho))
```

### Graphing outcomes

```{r}
# #define the categories I want to cycle through
# #unique() returns a vector, df, or array like x but with duplicate elements removed
# subjects <- unique(spearman_performance_summary$subject_id)
# onset_timing <- unique(spearman_performance_summary$date_vs_onset_illness)
# 
# #create and store the plots in a nested list like my data from earlier
# #expand.grid() creates a df from all combinations of supplied vectors 
# spearman_predictive_accuracy_plots <- expand.grid(subject = subjects, onset = onset_timing) %>% 
# #pmap maps in parallel 
#   pmap(function(subject, onset) {
#     spearman_performance_summary %>% 
#       filter(subject_id == subject, 
#              date_vs_onset_illness == onset,
#              pos_minus_neg > 0) %>% 
#       ggplot(aes(x = reorder(folder_name, significant_positive_taxa), 
#                  y = pos_minus_neg, 
#                  fill = median_rho)) +
#       #makes bar plot with actual data values (percent_significant_positive)
#       #default is stat = "counts"
#       geom_bar(stat = "identity", color = "black") +
#       #put folder names on the side for easier readability
#       coord_flip() +
#       scale_fill_gradient(low = "lightyellow",
#                           high = "darkgreen") +
#       labs(x = "Parameter Combination (folder_name)", 
#         y = "# Significant Positive Taxa - Significant Negative Taxa",
#         fill = "Median Rho",
#         title = paste("MICOM Parameter Rankings for ", subject),
#         subtitle = paste("Time relative to illness onset: ", onset,
#                          "\nRanked by # Significant Positive Taxa; color = Median Rho")
#       ) +
#       theme_minimal() +
#       theme(legend.position = "right")
#     
#   })
# 
# #name the plots
# names(spearman_predictive_accuracy_plots) <- expand.grid(
#   subject = subjects, onset = onset_timing) %>%
#   mutate(plot_name = paste(subject, onset, sep = "_")) %>%
#   pull(plot_name)
```

```{r}
# Define the unique subjects to iterate over
subjects <- unique(spearman_performance_summary$subject_id)

# Create and store the plots in a list using purrr::map
spearman_predictive_accuracy_plots <- purrr::map(subjects, function(subject) {
  spearman_performance_summary %>% 
    filter(subject_id == subject, 
           pos_minus_neg > 0) %>% 
    ggplot(aes(x = reorder(folder_name, pos_minus_neg), 
               y = pos_minus_neg, 
               fill = median_rho)) +
    geom_bar(stat = "identity", color = "black") +
    coord_flip() +
    scale_fill_gradient(low = "lightyellow",
                        high = "darkgreen") +
    labs(x = "Parameter Combination (folder_name)", 
         y = "# Significant Positive - # Significant Negative Taxa",
         fill = "Median Rho",
         title = paste("MICOM Parameter Rankings for", subject),
         subtitle = "Ranked by # Significant Positive - # Significant Negative Taxa
         \n color = Median Rho"
    ) +
    theme_minimal() +
    theme(legend.position = "right")
})

# Name the plots with the subject id
names(spearman_predictive_accuracy_plots) <- subjects
```

```{r}

# --- your existing labels ---
folder_labs <- c(
  "growth_M01_agora201_gurobi_wd_08"   = "AGORA2,WesternDiet,0.8", 
  "growth_M01_agora201_gurobi_wd_09"   = "AGORA2,WesternDiet,0.9", 
  "growth_M01_agora1_gurobi_vmhfat_08" = "AGORA1,HighFat/Prot+LowCarb,0.8",
  "growth_M01_agora1_gurobi_vmhfat_04" = "AGORA1,HighFat/Prot+LowCarb,0.4",
  "growth_M01_agora201_gurobi_vmhavg_09" = "AGORA2,AustrianAvg,0.9",
  "growth_M01_agora1_gurobi_vmhavg_07" = "AGORA1,AustrianAvg,0.7",
  "growth_M01_agora1_gurobi_vmhfiber_08" = "AGORA1,HighFiber,0.8",
  "growth_M01_agora1_gurobi_wd_09"       = "AGORA1,WesternDiet,0.9"
)

# Make sure nothing is grouped
spearman_performance_summary <- spearman_performance_summary %>% ungroup()

# Compute a total_taxa column if you have one; otherwise fall back to sig total
spearman_performance_summary <- spearman_performance_summary %>%
  mutate(
    total_taxa = if ("total_taxa" %in% names(.)) total_taxa
                 else (significant_positive_taxa + significant_negative_taxa),
    non_significant_taxa = pmax(
      total_taxa - (significant_positive_taxa + significant_negative_taxa), 0
    )
  )

# Long data with a 3rd "Non-significant" stratum
spearman_long <- spearman_performance_summary %>%
  select(subject_id, folder_name, pos_minus_neg,
         significant_positive_taxa, significant_negative_taxa, non_significant_taxa) %>%
  pivot_longer(
    c(significant_positive_taxa, significant_negative_taxa, non_significant_taxa),
    names_to = "sign", values_to = "count"
  ) %>%
  mutate(sign = recode(sign,
    significant_positive_taxa = "Positive",
    significant_negative_taxa = "Negative",
    non_significant_taxa     = "Non-significant"
  )) %>%
  # control stack order: Positive at base, then Negative, gray on top
  mutate(sign = factor(sign, levels = c("Non-significant","Negative","Positive"))) %>%
  ungroup()

subjects <- unique(spearman_long$subject_id)

spearman_predictive_accuracy_plots <- purrr::map(subjects, function(subject) {
  # Top 5 by # positive taxa
  top5 <- spearman_performance_summary %>%
    filter(subject_id == subject) %>%
    arrange(desc(significant_positive_taxa)) %>%
    slice_head(n = 5) %>%
    select(folder_name, significant_positive_taxa)

  # Keep only those folders
  df_sub <- spearman_long %>%
    filter(subject_id == subject, folder_name %in% top5$folder_name)

  # Order bars by # positive taxa (largest at top after coord_flip)
  ord_levels <- top5 %>% arrange(significant_positive_taxa) %>% pull(folder_name)
  df_sub <- df_sub %>% mutate(folder_name = factor(folder_name, levels = ord_levels))

  ggplot(df_sub, aes(x = folder_name, y = count, fill = sign)) +
    geom_col(color = "black") +
    coord_flip(clip = "off") +
    scale_x_discrete(labels = folder_labs) +
    scale_fill_manual(
      values = c(
        "Positive" = "#3b82f6",          # blue
        "Negative" = "#ef4444",          # red
        "Non-significant" = "grey80"     # gray fill to total taxa
      ),
      name = "Spearman Correlation"
    ) +
    labs(
      x = "Parameter Combo",
      y = "Total # Genera in Model",
      title = paste("Top 5 MICOM Parameter Combos for", subject)
    ) +
    theme_minimal(base_size = 18) +
    theme(
      legend.position = "right",
      plot.margin = margin(5.5, 25, 5.5, 5.5)
    )
})

names(spearman_predictive_accuracy_plots) <- subjects
```

```{r, fig.width = 9, fig.height=3}
spearman_predictive_accuracy_plots$M01
```

```{r}
M01_best_performing_condition <- spearman_results %>% 
  filter(folder_name == "growth_M01_agora201_gurobi_wd_08")
# 
# M01_best_pearson_condition <- pearson_results %>% 
#   filter(folder_name == "growth_M01_agora201_gurobi_wd_08")
```

```{r}
F01_best_performing_condition <- spearman_results %>% 
  filter(folder_name == "growth_F01_agora1_gurobi_wd_09")
```

```{r}
M02_best_performing_condition <- spearman_results %>% 
  filter(folder_name == "growth_M02_agora1_gurobi_vmhfat_07")
```

```{r}
# Define the unique subjects to iterate over
# subjects <- unique(pearson_performance_summary$subject_id)
# 
# # Create and store the plots in a list using purrr::map
# pearson_predictive_accuracy_plots <- purrr::map(subjects, function(subject) {
#   pearson_performance_summary %>% 
#     filter(subject_id == subject, 
#            pos_minus_neg > 0) %>% 
#     ggplot(aes(x = reorder(folder_name, pos_minus_neg), 
#                y = pos_minus_neg, 
#                fill = median_rho)) +
#     geom_bar(stat = "identity", color = "black") +
#     coord_flip() +
#     scale_fill_gradient(low = "lightyellow",
#                         high = "darkgreen") +
#     labs(x = "Parameter Combination (folder_name)", 
#          y = "# Significant Positive - # Significant Negative Taxa",
#          fill = "Median Rho",
#          title = paste("MICOM Parameter Rankings for", subject),
#          subtitle = "Ranked by # Significant Positive - # Significant Negative Taxa
#          \n color = Median Rho"
#     ) +
#     theme_minimal() +
#     theme(legend.position = "right")
# })
# 
# # Name the plots with the subject id
# names(pearson_predictive_accuracy_plots) <- subjects
```

```{r}
# pearson_predictive_accuracy_plots$M01
# pearson_predictive_accuracy_plots$F01
# pearson_predictive_accuracy_plots$M02
```

## EXTRA STUFF FOR PRESENTATION

```{r}
M01_2nd_best_performing_condition <- spearman_results %>% 
  filter(folder_name == "growth_M01_agora201_gurobi_wd_09")

M01_3rd_best_performing_condition <- spearman_results %>% 
  filter(folder_name == "growth_M01_agora201_gurobi_vmhavg_09")
```

```{r}
M01_best_filter <- sim_real_data %>% 
  filter(folder_name =="growth_M01_agora201_gurobi_wd_08")

```

```{r}
library(ggplot2)
library(dplyr)

# Filter for one genus
taxon_to_plot <- "Anaerostipes"

df_taxon <- M01_best_filter %>%
  filter(taxon == taxon_to_plot)

# Spearman correlation
cor_test <- cor.test(df_taxon$growth_rate, df_taxon$clr_change_abund, method = "spearman", exact = FALSE)

# Extract rho and p-value
rho <- round(cor_test$estimate, 2)
pval <- signif(cor_test$p.value, 2)

# Make the plot
ggplot(df_taxon, aes(x = growth_rate, y = clr_change_abund)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed", color = "steelblue") +
  labs(
    title = paste("Spearman ρ =", rho, "| p =", pval),
    x = "MICOM-predicted growth rate",
    y = "Change in CLR abundance"
  ) +
  theme_minimal()
```

```{r, fig.width=4, fig.height=3}
taxon_to_plot <- "Bacteroides"

df_taxon <- M01_best_filter %>%
  filter(taxon == taxon_to_plot)

cor_test <- cor.test(df_taxon$growth_rate, df_taxon$clr_change_abund, method = "spearman", exact = FALSE)

rho <- round(cor_test$estimate, 2)
pval <- signif(cor_test$p.value, 2)

ggplot(df_taxon, aes(x = growth_rate, y = clr_change_abund)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed", color = "steelblue") +
  labs(
    title = paste("Before illness | Spearman ρ =", rho, "| p =", pval),
    x = "MICOM-predicted growth rate",
    y = "Change in CLR abundance"
  ) +
  theme_minimal()
```

```{r}
library(ggplot2)
library(dplyr)
library(purrr)

# Filter to only "before" illness onset
df_before <- M01_best_filter 
#%>%
  #filter(date_vs_onset_illness == "before")

# Create output directory if it doesn't exist
if (!dir.exists("../images")) {
  dir.create("../images", recursive = TRUE)
}

# Function to create and save one plot per taxon
plot_and_save_taxon <- function(taxon_name) {
  df_taxon <- df_before %>% filter(taxon == taxon_name)

  if (nrow(df_taxon) < 3) return(NULL)  # Skip taxa with too few data points

  # Spearman correlation
  cor_test <- cor.test(df_taxon$growth_rate, df_taxon$clr_change_abund, method = "spearman", exact = FALSE)
  rho <- round(cor_test$estimate, 2)
  pval <- signif(cor_test$p.value, 2)

  # Create plot
  p <- ggplot(df_taxon, aes(x = growth_rate, y = clr_change_abund)) +
    geom_point(size = 2, alpha = 0.7) +
    geom_smooth(method = "lm", se = FALSE, linetype = "dashed", color = "steelblue") +
    labs(
      title = paste0("Taxon: ", taxon_name, "\nSpearman ρ = ", rho, ", p = ", pval),
      x = "MICOM-predicted growth rate",
      y = "Change in CLR abundance"
    ) +
    theme_minimal()

  # Save plot
  filename <- paste0("../images/", gsub("[ /]", "_", taxon_name), "_full_2.png")
  ggsave(filename, plot = p, width = 4, height = 2, dpi = 300)
}

# Get list of unique taxa
unique_taxa <- unique(df_before$taxon)

# Generate and save all plots
walk(unique_taxa, plot_and_save_taxon)
```
